visualFields 1.0.7: Tools for Analyzing the Field of Vision in \proglang{R}

### 2025-05-11 jss-article-rnw/article.Rnw
Started shortening Introduction.

The \pkg{visualFields} package, part of the OPI, is a collection of tools for analyzing the field of vision. It provides a framework for the development and use of innovative methods for visualization, statistical analysis, and clinical interpretation of visual-field loss and its change over time.

@@

The visual field of an eye is the extent of the world that can be seen. It refers to both the central and the peripheral vision. It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles, and those in the inferior and nasal being negative. In healthy eyes, the sensitivity to light contrast decreases as the visual angle increases (or as it is commonly said in perimetry, eccentricity). The difference in light sensitivity over the visual field as a function of eccentricity is known as Traquair's \emph{hill of vision} \citep{IMF:Traquair1938}. The figure below shows the hill of vision as rendered by Aulhorn \& Harms \citep{IMF:Aul1967}.

@@

\subsection{What is visual field testing?}

Visual field testing measures a patient's visual sensitivity at different locations of the visual field. Its purpose is to reconstruct as best as possible within the confines and limitations of a clinical setting the \textbf{hill of vision}. The most common type of visual field testing is \textbf{static automated perimetry (SAP)}, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field. In contrast, \textbf{campimetry} is performed with a flat screen. [CAMPIMETRY MAY BE USED MORE OFTEN IN THE FUTURE BECAUSE COMPUTER SCREENS/ VR DEVICES MAY BE USED FOR TELEMEDICINE; LEAVE IT?]

Usually, perimetry tests the ability to detect increments in stimulus luminance with respect to the background with an achromatic (not colored) stimulus. This is a form of perimetry known as \textbf{white-on-white perimetry}. However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity). The figure below shows the most common grid of locations tested during {SAP}. The gray ellipse corresponds to the blind spot of the hill of vision and, therefore, the red dots are expected to have no sensitivity.

@@

The visual fields package was created to analyze data from \textbf{SAP}. In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus. The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50\% of the times. Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.

There is another common form of perimetry, \textbf{kinetic perimetry}, where a light spot is moved from the periphery towards the center along a number of meridians. In this form of perimetry the location where the spot is first detected by the subject is noted and those points at different meridians are connected to find isopters of equal sensitivity on the retina. The statistical analysis for this form of perimetry is very different than for static perimetry and beyond the scope of this visualFields package.

\subsection{Performing visual field tests}
For the typical visual field test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere). The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination. In his hand, the observer has access to a clicker.

Observers now look at a fixation target (a dot, a cross, a diamond, or other) right in front of them in the center of the dome and are presented visual stimuli over their visual field. Observer are to press the clicker any time a stimulus appears anywhere in the dome These lights are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus. A press of the clicker will be linked to a presentation when it occurs within a given time window after the presentation. A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order. For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.

@@

\subsection{Psychophysical testing}
Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception. This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived.

The relationship between stimulus and perception is analyzed using mathematical model. Commonly, these models result in an estimate of a \textbf{threshold}, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.

Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog. Stimuli are not clearly seen when close to threshold. For this reason, patients sometimes feel that lengthy visual field testing is tedious and that it is as an unreliable examination. The latter is not clearly correct when fluctuations are taken into account and data are analyzed with appropriate statistical methods.

>>>>

The \pkg{visualFields} package, part of the OPI, is a collection of tools for analyzing the field of vision. It provides a framework for the development and use of innovative methods for statistical analysis, visualization, and clinical interpretation of visual-field loss and its change over time.

@@

The visual field is the extent of the world that can be seen and refers to both the central and the peripheral vision. It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles, and those in the inferior and nasal being negative.

In healthy eyes, the sensitivity to light contrast decreases with eccentricity (i.e. increasing visual angle. The difference in light sensitivity over the visual field as a function of eccentricity is known as Traquair's \emph{hill of vision} \citep{IMF:Traquair1938}. \ref{hill} below shows the hill of vision as rendered by Aulhorn \& Harms \citep{IMF:Aul1967}.

@@

\subsection{What is perimetry?}

Perimetry (visual field testing) measures a patient's sensitivity to light differences at different locations of the visual field. Thus, it is a psychometric way of quantifying the visual field. Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception. This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived. The relationship between stimulus and perception is analyzed using mathematical model. Commonly, these models result in an estimate of a \textbf{threshold}, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.

Its purpose is to estimate the \textbf{hill of vision} as best as possible within the confines and limitations of a clinical setting. The most common type of visual field testing is \textbf{static automated perimetry (SAP)}, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field. In contrast, \textbf{campimetry} is performed with a flat screen. Campimetry may be used more often in the future, because computer screens and virtual reality devices may be used for telemedicine.

The visual fields package was created to analyze data from standard automated perimetry\textbf{SAP}, where the light sensitivity is measured at fixed, discrete locations in the visual field. Kinetic perimetry, where locations of isosensitivity to fixed, discrete luminance increments, is beyond the scope of the visualFields package.

In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus. The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50\% of the times. Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.

For the typical SAP test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere). The perimeter is located in a dimly lit, calm room and the observer has been instructed by an experienced technician before the examination. During the test, the observer looks at a fixation target (a dot, a cross, a diamond, or other) in the center of the dome. The observer presses a button in his hand any time a stimulus appears anywhere in the dome. These stimuli are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus. A press of the button will only be linked to a presentation when it occurs within a given time window after the presentation. A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order. For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.

The figure below shows the most common grid of locations tested during {SAP}. The gray ellipse corresponds to the blind spot of the hill of vision and, therefore, the red dots are expected to have no sensitivity.

@@

@@
